---
title: 'Using functions with Data'
author: "Naomi Tague"
date: "January, 2020"
output:
  slidy_presentation:
   highlight: pygments
  html_document: default
  pdf_document: default
  ioslides_presentation:
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)
knitr::opts_chunk$set(error=TRUE)
library(tidyverse)
```

# A reminder about naming objects in R

Note that by *sourcing* a function - it will essentially overwrite anything else in your workspace with the same name 

*Warning* - if you name your function, the same as an internal R function, your new function will take precidence, and **hide** the internal R function

In R, functions are organized in **Packages**

You've probably loaded different packages, that provide different functions
There are a number of packages **base**, **stats** that are automatically loaded
You can usually find the package associated with any function from online help

* consider **runif** function in the **stats** package

[runif](https://stat.ethz.ch/R-manual/R-patched/library/stats/html/Uniform.html)

To use a function associated with a particular package

**package::function**


# Generating Data for your function

You often want to apply your function to a range of data, that can come from

* files you read in to R (measured height and flow rate values)
* output from other functions/models 
* data that you generate
  * sensitivity analysis
  * testing your model
  * stochastic models


# Random Numbers as Inputs

* sample from distributions (normal, uniform, poisson), which distribution depends on the model
* R has many tools for generating samples from distributions with known parameters (such as mean, standard deviation, or min/max)
  *  generating rainfall for a hydrologic model given know mean and variance of rainfall
  
* R also has tools for picking samples from collections 
  * generating fish catches based on populations for an economic model
  
  
  Others?
***

# Steps for running your model over multiple inputs

1. design a data structure to store results: sometimes this is automatic but not always
2. generate the input data
3. apply to the model



# Lets consider a new function estimates the power required to keep a car moving at a given speed



<span style="color:coral">
**Pb = crolling * m * g * V + 1/2 A * pair * cdrag * V3**
</span>

where *crolling* and *cdrag* are rolling and aerodynamic resistive coefficients, typical values are 0.015 and 0.3, respectively

*V*: is vehicle speed (assuming no headwind) in m/s (or mps) m: is vehicle mass in kg

*A* is surface area of car (m2)

*g* is acceleration due to gravity (9.8 m/s2) pair = density of air (1.2kg/m3)

*Pb* is power in Watts


# Code

```{r powerexample}

#' Power Required by Speed
#'
#' This function determines the power required to keep a vehicle moving at 
#' a given speed
#' @param cdrag coefficient due to drag default=0.3 
#' @param crolling coefficient due to rolling/friction default=0.015
#' @param v vehicle speed (m/2)
#' @param m vehicle mass (kg)
#' @param A area of front of vehicle (m2)
#' @param g acceleration due to gravity (m/s) default=9.8
#' @param pair (kg/m3) default =1.2
#' @return power (W)

autopower = function(V, m, A, cdrag=0.3, crolling=0.015,pair=1.2,g=9.8) {
  P = crolling*m*g*V + 1/2*A*pair*cdrag*V**3
  return(P)
}

```

I've also stored the autopower.R function as a file = using the recommended naming convention

**autopower.R**


# Generating data for the function

Example use: Imagine we want to see how much power is generated given scenario where we know the mean and standard deviation of vehicles speeds


```{r sampling}

#source("./src/dataf/R/autopower.R")

# generate sample speeds from a distribution
nsample = 100
speeds = rnorm(mean=25, sd=4, n=nsample)

# Step 1  create data frame to store results 
# how many simulations, what do you want to keep

#create a dataframe that has rows for each model run
# columns for height, flowrate and power estimate
results = data.frame(speed=speeds, power=NA)

head(results)

# if you only have one input and everything elese is the same, R is smart enough
# to create a set of outputs
results$power = autopower(V=speeds, A=25, m=20000)

# ALWAYS look at your results to make sure it make sense
ggplot(results, aes(speed, power/1000))+geom_point()+labs(x="speed in m/s", y="power in kW")
ggplot(results, aes(x="Across Speed",y=power/1000))+geom_boxplot()+labs(y="power in kW")

```


# Multiple inputs 

For a function with more than one input that could have multiple values, R can handle this
if 

* all (multiple) inputs have the same number of values 
* all other inputs or paraameters have a single value


For other cases, additional instructions would need to be included in the function to help R understand what to do


Consider our reservoir function  (see Rmd from functions2 for power_gen.R code)

```{r sampling2}

#source("src/dataf/R/power_gen.R")

power_gen = function(height, flow, rho=1000, g=9.8, Keff=0.8) {
  # make sure inputs are positive and if not set as NA so result will be NA
  height = ifelse( (height < 0), NA, height)
  
  # an alterative is to exit the function prematurely 
  flow = ifelse((flow < 0), return("flow must be greater than zero"), flow)
  
  #    return("Height cannot be less than 0")
  
  # calculate power
  result = rho * height * flow * g * Keff
  return(result)
}

# single
power_gen(height=2, flow=2)

# multiple heights
# generate a strucutre to store results
# consider all heights from 1 to 100

resv_results = data.frame(height=seq(from=1, to=100))

# compute power generated for EACH height
resv_results$power = power_gen(resv_results$height, flow=2)

# plot
ggplot(resv_results, aes(height, power))+geom_point(size=3)+labs(y="Power (KW)", x="Height (m)")

# random combinations of height and flow
resv_results$height = rnorm(mean=10, sd=2, n=100)
resv_results$flow = rnorm(mean=20, sd=1, n=100)
# compute power generated for EACH height
resv_results$power = power_gen(resv_results$height, flow=resv_results$flow)

# plot
ggplot(resv_results, aes(height, power, col=flow))+geom_point(size=3)+labs(y="Power (KW)", x="Height (m)")

# Unequal lengths - this will "work" but inputs get recycled unclear what is happening
# NOT recommended

eheight = seq(from=1, to=50)
eflow = runif(min=0.1, max=2, n=100)
resv_result = power_gen(eheight, eflow)

# better
eheight = seq(from=1, to=50)
eflow = runif(min=0.1, max=2, n=50)
resv_result = power_gen(eheight, eflow)

# another example of the problem of unequal length inputs
power_gen(eheight, eflow, g=c(9,10,11))

#better
g = runif(min=8, max=10, n=50)
resv_result = power_gen(eheight, eflow, g)


```
# Assignment 1


With a partner

Write function to compute something that is of interest to you, make sure it has at least 2 inputs and several parameters

Add some error checking (at least two different kinds)

Save your function in a .R file

(Make sure you add comments to your function)

In an Rmarkdown document, generate saample data for your function in two different ways, 

Apply the data to your function

Graph and add some text describing how to interpret the results

Upload both the function and the Rmarkdown file to your github - in a subfolder called Assignment 1

Add a link to your github site for Assignment 1 on Gauchospace

