---
title: 'Functions Day 2'
author: "Naomi Tague"
date: "January, 2020"
output:
  slidy_presentation:
   highlight: pygments
  html_document: default
  pdf_document: default
  ioslides_presentation:
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)
knitr::opts_chunk$set(error=TRUE)
library(tidyverse)
```


## STEPS IN Program Design

1. Clearly define your goal in words/figures as precisely as possible what do you want your program to do
      + modules
      + for each module,  inputs and parameters
      + for each module, outputs
    

2. Implement AND document as you go
3. Test
4. Refine
5. Test again
6. Create User Documentation and Distribute

---

## Functions in R

Format for a basic function in R

<span style="color:coral">
\#' documentation that describes inputs, outputs and what the function does
FUNCTION NAME = function(inputs, parameters) {
body of the function (manipulation of inputs)
return(values to return)
}
</span>

In R, inputs and parameters are treated the same; but it is useful to think about them separately in designing the model - collectively they are sometimes referred to as arguments

ALWAYS USE Meaningful names for your function, its parameters and variables calculated within the function

---

## A simple function example

Contract

*Input*: Reservoir height (m) and flow rate (m3/sec) (as a single value in time or a time series)

*Output*: Instantaneous power generation (W/s)  (as a single value or value for each reservoir height and flow rate combination)

*Parameters*: KEfficiency, ρ (density of water), g (acceleration due to gravity)

How it wil work - *body* of the function 

P = ρ * h * r * g * KEff;

P is Power in watts, ρ is the density of water (~1000 kg/m3), h is height in meters, r is flow rate in cubic meters per second, g is acceleration due to gravity of 9.8 m/s2, KEff is a coefficient of efficiency ranging from 0 to 1.

---
## Implementation

* In R we combine inputs/parameters in the first line of the function definition

* we can provide "default" values that can be overwritten by the function user

* Body includes all calculations between { and }
  + anything inside the body is not "seen" by the workspace so you can re-use variables
  
* *return* tells R what the output is


```{r eval=TRUE, echo=TRUE  }

# function definition
power_gen = function(height, flow, rho=1000, g=9.8, Keff=0.8) {
result = rho * height * flow * g * Keff
return(result)
}
```

---
# Using the function and scoping

To use the function we substitute "actual" values for each argument (parameter and inputs)
(unless there are defaults)

R assumes that arguments are in the order that they were specified in the function definition - unless
you refer to them by name
```{r eval=TRUE, echo=TRUE  }

power_gen(20,1)

power_gen(height=20, flow=1)

power_gen(flow=1, height=20)

# change the defaults
# what if efficiencies is less

power_gen(20,1, Keff=0.2)

```

---
# Scoping


The scope of a variable in a program defines where it can be “seen” Variables defined inside a function cannot be “seen” outside of that function

There are advantages to this - the interior of the building block does not ‘interfere’ with other parts of the program

See below

```{r eval=TRUE, echo=TRUE  }

# specifying function name without () shows code
# try it for internal R functions
power_gen

# use the function and save the results
reservoir1 = power_gen(height=4, flow=7)

reservoir2 = power_gen(height=4, flow=7, Keff=0.2)

# note that internal variables are not defined
height
Keff
results

# only the outputs
reservoir1
reservoir2


```

---
# Some conventions (helpful later in the course)

* Always write your function in a text editor and then copy into R

* By convention we name files with functions in them by the name of the function.R
e.g. **power_gen.R**

* you can have R read a text file by source(“power_gen.R”) - make sure you are in the right working directory

* keep organized by keeping all functions in a subdirectory called *R*

*  Eventually we will want our function to be part of a package (a library of many functions) - to create a package you must use this convention (name.R) place all function in a directory called “R”


Make your own power generation function - add some complexity (cut off height for power generation; maximum capacity?)


# Error Checking

What will your function do if user gives you garbage data, imagine that computes microbial growth rates with temperature

* Two options

<span style="color:skyblue">
1. error-checking (nice)
</span>

 if temperature < -100 or > 100
 
*  set results to NA
*  or exit function with a warming
 
<span style="color:skyblue"> 
2. assume user reads the contract :) return unrealistic values
</span>

if input -999.0, will still try to output growth rate
 
 
 Error-checking is helpful if you are going to build a model made up of many functions- why?
 

# How to add error checking

Consider this example

```{r eval=TRUE, echo=TRUE  }

power_gen = function(height, flow, rho=1000, g=9.8, Keff=0.8) {
  # make sure inputs are positive and if not set as NA so result will be NA
   if (height < 0) height=NA
  
 # an alterative is to exit the function prematurely 
  if (flow < 0) return("Flow cannot be less than 0")
  #    return("Height cannot be less than 0")
      
  # calculate power
    result = rho * height * flow * g * Keff
  return(result)
}

power_gen(2,3)
power_gen(-2,3)
power_gen(2,-3)
```


While this works for a case with one height and one flow
It will have problems if we use multiple heights - (a later topic)

A better implementation if you think the function might be run for multiple heights

```{r eval=TRUE, echo=TRUE  }

power_gen = function(height, flow, rho=1000, g=9.8, Keff=0.8) {
  # make sure inputs are positive and if not set as NA so result will be NA
   height = ifelse( (height < 0), NA, height)
  
 # an alterative is to exit the function prematurely 
  flow = ifelse((flow < 0), return("flow must be greater than zero"), flow)

  #    return("Height cannot be less than 0")
      
  # calculate power
    result = rho * height * flow * g * Keff
  return(result)
}

power_gen(2,3)
power_gen(-2,3)
power_gen(2,-3)
```